---
title: "PC3: Project Proposal + Data"
author: Anne Johnson, Carly Waldeck, David Roth, Joseph Piini
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(knitr)

gdp <- read.csv("gdp_pcap.csv")
food_supply <- read.csv("food_supply_kilocalories_per_person_and_day.csv")
```

## Introduction

**Research Question**

Over the course of this project, our groups hopes to answer the following question:

*Is there an association between gross domestic product (per capita) and food supply?*

**Summary of Data**

Our team will explore two datasets in our analysis: Gross domestic product (GDP) per capita and food supply.

GDP per capita provides a measure of a country's total economic output, divided by the nation's population. The variable is adjusted for inflation by fixing prices to 2017 dollars, and measured in Purchasing Power Parity (PPP) dollars. PPP dollars offer a virtual currency used to facilitate international economic comparisons. The [dataset](https://www.gapminder.org/data/documentation/gd001/) contains observations for 195 countries, providing estimates spanning from 1800 to 2100. The dataset from 1820 to 1990 is from the Maddison Project Database (https://www.rug.nl/ggdc/historicaldevelopment/maddison/releases/maddison-project-database-2023) and the data from 1990 to 2022 is from the World Bank(https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD). When the World Bank or the Maddison Project did not have data, the data was taken from estimates from the Penn World Table. Then, if data is still needed Gapminder uses previous versions of data. (https://www.gapminder.org/data/documentation/gd001/#v26)

The [food supply dataset](https://www.fao.org/faostat/en/#home), sourced from the United Nation's Food and Agriculture Organization, provides insight into a country's food security and nutritional status. The dataset offers annual observations of a country's food supply by measuring kilocalories per person per day. The dataset contains observations for 179 countries, spanning from 1961 to 2018.

## Data Cleaning/Joining

```{r data_cleaning}
# Joining datasets to include one obs. per country/year combination

gdp_longer <- gdp |>
  mutate(across(.cols = !country,
                .fns = ~ case_when(str_detect(.x, "k") ~ as.numeric(str_remove(.x, "k")) * 1000, TRUE ~ as.numeric(.x)))) |>
  pivot_longer(cols = !country,
               names_to = "year")

food_supply_longer <- food_supply |>
  mutate(across(.cols = !country,
                .fns = ~as.numeric(.x))) |>
  pivot_longer(cols = !country,
               names_to = "year")

data_joined <- gdp_longer |>
  inner_join(food_supply_longer, # only include matching rows
             join_by(country, year)) |>
  rename(gdp_pcap = "value.x",
         food_supply = "value.y") |>
  drop_na() |>
  mutate(year = as_factor(str_extract(year, pattern = "[:digit:]{1,}")),
         country = as_factor(country))

```

## Discussion

**Hypothesized Relationship**

We hypothesize that there will be a positive linear relationship with GDP and calories. Foremost, there is a positive linear association with GDP per capita and quality of life. Using this knowledge, that as GDP per capita increases, food abundance would increase, causing the number of calories consumed to also increase. Also, using GDP per capita as a measure of income, as income rises, one is able to afford more. Under the assumption of monotonicity, more is better, and the income effect, one would believe that an increase in overall income would allow people to buy more food (more is better) and have to work less (more time to leisure--eat).(Sands 2024). Furthermore, poorer nations, nations with less GDP per capita suffer more from malnutrition. 

#https://www.worldbank.org/en/topic/nutrition/overview

**Data Cleaning Process**

The food supply data contains sixteen fewer countries than the GDP data, so our analysis may be limited by the removal of these countries. Additionally, the GDP dataset contains data for a much wider range of years (1800-2100) than the food supply dataset (1961-2018), and there are several observations within the food supply dataset that are missing values. Our data was cleaned to only include years and countries where both sets of data are available.

After the data cleaning process, our dataset to be used moving forward contains a GDP per capita and food supply value for each observational unit (year/country combination). Our cleaned data includes 170 countries, spanning from 1961 to 2018.

*Discuss any data cleaning process and decisions that may impact the analysis.*

The food supply data contains sixteen more countries than the GDP data, so our analysis may be limited by the removal of these countries. Additionally, the GDP dataset contains data for a much wider range of years (1800-2100) than the food supply dataset (1961-2018), and there are several observations within the food supply dataset that are missing values. Our data was cleaned to only include years and countries where both sets of data are available.

After the data cleaning process, our dataset to be used moving forward contains a GDP per capita and food supply value for each observational unit (year/country combination). Our cleaned data includes 170 countries, spanning from 1961 to 2018.
>>>>>>> c0b26bb39f33a4ee26806744feb05369bbf8c0bb
>>>>>>> c2fb5577daa1d38a0e45eae4e0e3067621518b61
>>>>>>> 291f43f00accb9b580115bbfb2ca7684eccffe2d

## References

\[1\] GapMinder, “GD001,” *Gapminder.org*, 2011. https://www.gapminder.org/data/documentation/gd001/

\[2\] Food and Agriculture Organization of the United Nations, “FAOSTAT,” *www.fao.org*, 2024. https://www.fao.org/faostat/en/#home

\[3\] R. Sands, “Patterns of global food consumption expected to shift in next quarter century as population, incomes rise | Economic Research Service,” Jun. 20, 2024. https://www.ers.usda.gov/amber-waves/2024/june/patterns-of-global-food-consumption-expected-to-shift-in-next-quarter-century-as-population-incomes-rise (accessed May 23, 2025).

\[4\] “Overview,” World Bank. https://www.worldbank.org/en/topic/nutrition/overview
