---
title: "PC3: Project Proposal + Data"
author: Anne Johnson, Carly Waldeck, David Roth, Joseph Piini
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)

gdp <- read.csv("gdp_pcap.csv")
food_supply <- read.csv("food_supply_kilocalories_per_person_and_day.csv")
```

## Introduction

**Proposal**

TBD

**Summary of Data**

TBD

## Data Cleaning

```{r data_cleaning}
# Joining datasets to include one obs. per country/year combination

gdp_longer <- gdp |>
  mutate(across(.cols = !country,
                .fns = ~as.numeric(.x))) |>
  pivot_longer(cols = !country,
               names_to = "year")

food_supply_longer <- food_supply |>
  mutate(across(.cols = !country,
                .fns = ~as.numeric(.x))) |>
  pivot_longer(cols = !country,
               names_to = "year")

data_joined <- gdp_longer |>
  inner_join(food_supply_longer, # only include matching rows
             join_by(country, year)) |>
  rename(gdp_pcap = "value.x",
         food_supply = "value.y") |>
  drop_na() |>
  mutate(year = as_factor(str_extract(year, pattern = "[:digit:]{1,}")),
         country = as_factor(country))

```

## Discussion

**Hypothesized Relationship**

*Provide an explanation for why you think they would have that relationship (including at least one outside source).*

**Data Cleaning Process**

*Discuss any data cleaning process and decisions that may impact the analysis.*

## References

\[1\] GapMinder, “GD001,” *Gapminder.org*, 2011. https://www.gapminder.org/data/documentation/gd001/

\[2\] Food and Agriculture Organization of the United Nations, “FAOSTAT,” *www.fao.org*, 2024. https://www.fao.org/faostat/en/#home

‌
