---
title: "PC4: Linear Regression"
author: "Anne Johnson, Carly Waldeck, David Roth, Joseph Piini"
format: html
editor: visual
warning: FALSE
---

## Setup

```{r}
library(tidyverse)
library(gt)
library(kableExtra)
```

```{r}
# load in data
food <- read_csv("food_supply_kilocalories_per_person_and_day.csv")
gdp <- read_csv("gdp_pcap.csv")
```

## 2.1 Data Visualization

## 2.2 Linear Regression

```{r}
# clean data then take mean of data from 2012-2018 so we have one x and one y value for each country

food_summary <- food |>
  select("country", "2012", "2013", "2014", "2015",
         "2016", "2017", "2018") |>
  mutate(avg_food = rowMeans(across(`2012`:`2018`), na.rm = TRUE))|> 
  drop_na() |>
  select(country, avg_food)

gdp_summary <- gdp |>
  select(country, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`) |>
  mutate(across(`2012`:`2018`, ~ {
    ifelse(str_detect(.x, "k$"),
           as.numeric(str_remove(.x, "k")) * 1000,
           as.numeric(.x)
    )
  })) |>
  mutate(average_gdp = rowMeans(across(`2012`:`2018`), na.rm = TRUE)) |>
  select(country, average_gdp)

food_gdp <- left_join(food_summary, gdp_summary, by = "country")
```

```{r}
## run linear regression
model <- lm(average_gdp ~ avg_food, data = food_gdp)
summary(model)
```

```{r}
## scatterplot with line of best fit
ggplot(data = food_gdp, aes(x = avg_food, y = average_gdp)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Average Food vs Average GDP",
    x = "Average Food Supply",
    y = "Average GDP"
  ) +
  theme_minimal()
```

## 2.3 Model Fit

```{r}

# find varience
A <- var(food_gdp$average_gdp, na.rm = TRUE)     
B <- var(fitted(model), na.rm = TRUE)                
resid_var <- var(residuals(model), na.rm = TRUE)      
r_squared <- B / A 

# create table
tibble(
  Metric = c(
    "Variance in response",
    "Variance in fitted values",
    "Variance in residuals",
    "Model R²"
  ),
  Value = c(A, B, resid_var, r_squared)
) |>
  mutate(Value = round(Value, 2)) |>
  kable(format = "html", caption = "Regression Summary Statistics") |>
  kable_styling(bootstrap_options = c("hover", "condensed",
                                      "responsive"), full_width = FALSE,
                position = "left")

```

Our regression model accounts for 51% of the variability in the response. This means that 49% of the variability in GDP remains unexplained, suggesting that while the model captures some of teh relationship, it does not provide a strong fit. Ideally the R\^2 value would be much higher, to indicate a more reliable and predictive model.
