---
title: "PC4: Linear Regression"
author: "Anne Johnson, Carly Waldeck, David Roth, Joseph Piini"
format: html
editor: visual
embed-resources: true
warning: FALSE
---

## Setup

```{r}
library(tidyverse)
library(gganimate)
library(gt)
library(kableExtra)
library(gifski)
```

```{r}
# load in data
food <- read_csv("food_supply_kilocalories_per_person_and_day.csv")
gdp <- read_csv("gdp_pcap.csv")
```

## 2.1 Data Visualization

```{r}
# clean data then take mean of data from 2012-2018 so we have one x and one y value for each country

food_summary <- food |>
  select("country", "2012", "2013", "2014", "2015",
         "2016", "2017", "2018") |>
  mutate(avg_food = rowMeans(across(`2012`:`2018`), na.rm = TRUE))|> 
  drop_na() |>
  select(country, avg_food)

gdp_summary <- gdp |>
  select(country, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`) |>
  mutate(across(`2012`:`2018`, ~ {
    ifelse(str_detect(.x, "k$"),
           as.numeric(str_remove(.x, "k")) * 1000,
           as.numeric(.x)
    )
  })) |>
  mutate(average_gdp = rowMeans(across(`2012`:`2018`), na.rm = TRUE)) |>
  select(country, average_gdp)

food_gdp <- left_join(food_summary, gdp_summary, by = "country")
```

```{r}
## scatterplot with line of best fit
ggplot(data = food_gdp, aes(x = log(average_gdp), y = log(avg_food))) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Average Food vs Average GDP",
    x = "Average GDP",
    y = "Average Food Supply"
  ) +
  theme_minimal()
```

```{r}
# Joining datasets to include one obs. per country/year combination

gdp_longer <- gdp |>
  mutate(across(.cols = !country,
                .fns = ~ case_when(str_detect(.x, "k") ~ as.numeric(str_remove(.x, "k")) * 1000, TRUE ~ as.numeric(.x)))) |>
  pivot_longer(cols = !country,
               names_to = "year")

food_supply_longer <- food |>
  mutate(across(.cols = !country,
                .fns = ~as.numeric(.x))) |>
  pivot_longer(cols = !country,
               names_to = "year")

data_joined <- gdp_longer |>
  inner_join(food_supply_longer, # only include matching rows
             join_by(country, year)) |>
  rename(gdp_pcap = "value.x",
         food_supply = "value.y") |>
  drop_na() |>
  mutate(year = as_factor(str_extract(year, pattern = "[:digit:]{1,}")),
         country = as_factor(country))

```

```{r}
#| eval: false

# create animated scatterplot

data_joined_2 <- data_joined |>
  mutate(year = as.numeric(as.character(year)))
animation <- ggplot(data_joined_2, aes(x = log(gdp_pcap), y = log(food_supply), color = country)) +
  geom_point(size = 3, show.legend = FALSE) +
  labs(
    title = "GDP vs Food Supply Over Time",
    subtitle = "Year: {frame_time}",
    x = "GDP per Capita (USD)",
    y = "Food Supply (kcal/person/day)"
  ) +
  theme_minimal() +
  transition_time(year) +
  ease_aes('linear')

animate(animation, width = 800, height = 600, nframes = 58, fps = 10)

```

## 2.2 Linear Regression

```{r}
## run linear regression
model <- lm(log(avg_food) ~ log(average_gdp), data = food_gdp)
summary(model)
```

The following equation represents the fitted regression model, where $\widehat{food_i}$ represents the predicted average food supply (kcal/person/day) and $GDP_i$ represents average GDP per capita (USD).

$$\log{(\widehat{food}_i)} = 6.895 + 0.1125\log{(GDP_i)}$$

A 1% increase in average GDP per capita is associated with approximately a 0.1125% increase in average food supply.

The expected natural logarithm of average food supply is 6.8950 when the natural logarithm of average GDP per capita is zero.

## 2.3 Model Fit

```{r}

# find variance
A <- var(log(food_gdp$avg_food), na.rm = TRUE)     
B <- var(fitted(model), na.rm = TRUE)                
resid_var <- var(residuals(model), na.rm = TRUE)      
r_squared <- B / A 

# create table
tibble(
  Metric = c(
    "Variance in response",
    "Variance in fitted values",
    "Variance in residuals",
    "Model R²"
  ),
  Value = c(A, B, resid_var, r_squared)
) |>
  mutate(Value = round(Value, 2)) |>
  kable(format = "html", caption = "Regression Summary Statistics") |>
  kable_styling(bootstrap_options = c("hover", "condensed",
                                      "responsive"), full_width = FALSE,
                position = "left")

```

Our regression model accounts for 62% of the variability in the response. This means that 38% of the variability in GDP remains unexplained, suggesting that while the model captures some of the relationship, it does not provide a strong fit. Ideally the R\^2 value would be much higher, to indicate a more reliable and predictive model.
